Code Repository

### 2025-06-16 Code Repository.Rmd
updates

Plotting theoretical null distribution of summed test statistic composite
```{r}
ggplot(data = null_normal_distribution.df, aes(x =null_normal_distribution )) + geom_histogram(bins =250, colour = "skyblue", fill = "skyblue") + labs(title = "Normal Null Distribution", x = "", y = "Frequency") +
  theme_minimal() +   geom_vline(xintercept = quantile(null_normal_distribution, probs = c(0.025, 0.975)), color = "black", linetype = "dashed", linewidth = 0.75) + 
  geom_vline( xintercept = study_summed_composite, color = "gold", size = 1 ) + 
  annotate("text", x = study_summed_composite, y = 10000, label = "EMERGE Composite", color = "gold", angle = 90, vjust = -0.5)
```

@@

```
Plotting theoretical null distribution of squared test statistic composite with eigenvalue weights
```{r}
ggplot(data = theoretical_null_squared, aes(x =chi_sq )) + geom_histogram(bins =250, colour = "lightsalmon", fill = "lightsalmon") + labs(title = "", x = "", y = "") +
  theme_minimal() + 
  geom_vline(xintercept = quantile(theoretical_null_squared$chi_sq, probs = c(0.95)), color = "red", linetype = "dashed", linewidth = 0.75) +
  annotate("text", x = 8.2, y = 25000, label = "95th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5) +
  theme(panel.grid.minor = element_blank())+
  geom_vline( xintercept = study_squared_composite, color = "steelblue", size = 1 ) + 
  annotate("text", x = study_squared_composite, y = 25000, label = "Composite", color = "steelblue", angle = 90, vjust = -0.5, size = 5)

@@

Reprocessing tmpDat into the suitable format "inverted_data" for the LatVarfunc
Requires setting columns in a specific order and inverting the data
Column 1 is patient ID
Column 2 is treatment (binary 0,1)
Column 3 is gw32
Column 4 is gw38
Column 5 is Insulin (binary 0, 1)
Make sure to remove ## for the code to run
```{r}
data <- tmpDat[,c("Label", "binary_group", "gw32", "gw38", "binary_insulin")]
inverted_data <- data[complete.cases(data), ]
inverted_data$binary_insulin <- ifelse(inverted_data$binary_insulin == 1, 0, 1)
inverted_data$gw32 <- inverted_data$gw32 * -1
inverted_data$gw38 <- inverted_data$gw38 * -1
inverted_eta <- c(-5.1, -5.1) ## these are the "cut-off" points for the continuous variables
results <- LatVarfunc(inverted_data, inverted_eta)
names(results) <- c("logOR_LCI", "logOR", "logOR_UCI")
```

Power estimation of Latent variable analysis
*Note that this takes a long time to run, and as such we only ran it for two sample sizes, 100 and 150*
```{r}
set.seed(24253739)
power_function_null_dist <- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {
    placebo_group <- inverted_data[inverted_data$binary_group == 0, ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 100
##null_dist_summed_TS_results <- sapply(sample_vector, function(n_value) {power_function_null_dist(n_value)})
power_100_latvar<-0.99
power_150_latvar <- 0.99
```

@@

Applying power estimation function
```{r}
sample_vector <- seq(from = 30, to = 540, by = 10)
unified_power_results <- sapply(sample_vector, function(n_value) {
  power_function("gw32", "gw38", "binary_insulin", "binary_EMERGE", n_value)
})
unified_power_results_df <-unified_power_results
unified_power_results_df <- as.data.frame(t(unified_power_results)) 
colnames(unified_power_results_df) <- c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS")
unified_power_results_df$sample_size <- sample_vector

>>>>

Plotting theoretical null distribution of summed test statistic composite
```{r}
ggplot(data = null_normal_distribution.df, aes(x =null_normal_distribution )) + geom_histogram(bins =250, colour = "darksalmon", fill = "darksalmon") + labs(x = "Summed Composite Test Statistic (SC)", y = "Frequency") +
  theme_minimal() +   geom_vline(xintercept = quantile(null_normal_distribution.df$null_normal_distribution, probs = c(0.025, 0.975)), color = "red", linetype = "dashed", linewidth = 0.75) + 
  annotate("text", x = quantile(null_normal_distribution.df$null_normal_distribution, 0.025), y = 12000, label = "2.5th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5)+
  annotate("text", x = quantile(null_normal_distribution.df$null_normal_distribution, 0.975), y = 12000, label = "97.5th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5)+
  geom_vline( xintercept = study_summed_composite, color = "steelblue", size = 1 ) + 
  annotate("text", x = study_summed_composite, y = 10000, label = "EMERGE (SC = -9.1)", color = "steelblue", angle = 90, vjust = -0.5)
```

@@

```
Plotting theoretical null distribution of squared test statistic composite with eigenvalue weights
```{r}
ggplot(data = theoretical_null_squared, aes(x =chi_sq )) + geom_histogram(bins =250, colour = "lightsalmon", fill = "lightsalmon") + labs(x = "Squared Composite Test Statistic (QC)", y = "Frequency") +
  theme_minimal() + 
  geom_vline(xintercept = quantile(theoretical_null_squared$chi_sq, probs = c(0.95)), color = "red", linetype = "dashed", linewidth = 0.75) +
  annotate("text", x = quantile(theoretical_null_squared$chi_sq, 0.95), y = 25000, label = "95th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5) +
  theme(panel.grid.minor = element_blank())+
  geom_vline( xintercept = study_squared_composite, color = "steelblue", size = 1 ) + 
  annotate("text", x = study_squared_composite, y = 25000, label = "Composite (QC = 29.2)", color = "steelblue", angle = 90, vjust = -0.5, size = 5)
```

@@

Reprocessing tmpDat into the suitable format "inverted_data" for the LatVarfunc
Requires setting columns in a specific order and inverting the data
Column 1 is patient ID
Column 2 is treatment (binary 0,1)
Column 3 is gw32
Column 4 is gw38
Column 5 is Insulin (binary 0, 1)
Make sure to remove ## for the code to run
```{r}
data <- tmpDat[,c("Label", "binary_group", "gw32", "gw38", "binary_insulin")]
inverted_data <- data[complete.cases(data), ]
inverted_data$binary_insulin <- ifelse(inverted_data$binary_insulin == 1, 0, 1)
inverted_data$gw32 <- inverted_data$gw32 * -1
inverted_data$gw38 <- inverted_data$gw38 * -1
inverted_eta <- c(-5.1, -5.1) ## these are the "cut-off" points for the continuous variables
results_lat_var <- LatVarfunc(inverted_data, inverted_eta)
names(results_lat_var) <- c("logOR_LCI", "logOR", "logOR_UCI", "logRR_LCI", "logRR", "logRR_UCI", "RD_LCI", "RD", "RD_UCI",
  "Prob_Treat", "Prob_Control", "alpha0", "alpha1", "beta0", "beta1", "psi0", "psi1", "log_sig1", "log_sig2",
  "logit_rho12", "logit_rho13", "logit_rho23", "se_alpha0", "se_alpha1", "se_beta0", "se_beta1", "se_psi0", "se_psi1",
  "se_log_sig1", "se_log_sig2", "se_logit_rho12", "se_logit_rho13", "se_logit_rho23")
results_lat_var_df <- as.data.frame(t(results_lat_var))

log_odds_ratio    <- round(as.numeric(results_lat_var_df[1, 1:3]), 3)
log_risk_ratio    <- round(as.numeric(results_lat_var_df[1, 4:6]), 3)
risk_difference   <- round(as.numeric(results_lat_var_df[1, 7:9]), 3)

lat_var_results_table <- data.frame(Measure = c("Log Odds Ratio", "Log Risk Ratio", "Risk Difference"),
  "Lower Confidence Interval" = c(log_odds_ratio[1], log_risk_ratio[1], risk_difference[1]),
  `Point Estimate` = c(log_odds_ratio[2], log_risk_ratio[2], risk_difference[2]),
  `Upper Confidence Interval` = c(log_odds_ratio[3], log_risk_ratio[3], risk_difference[3])
)
write.csv(lat_var_results_table, "Latent variable results table.csv", row.names = FALSE)
```

Power estimation of Latent variable analysis
*Note that this takes a long time to run, and as such we only ran it for two sample sizes, 100 and 150*
```{r}
set.seed(24253739)
lat_var_power_function<- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {
    placebo_group <- inverted_data[inverted_data$binary_group == 0, ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
lat_var_power_results <- sapply(sample_vector, function(n_value) {lat_var_power_function(n_value)})
power_100_latvar<-0.83
power_200_latvar <- 0.97
```

@@

Applying power estimation function
```{r}
sample_vector <- seq(from = 30, to = 540, by = 10)
unified_power_results <- sapply(sample_vector, function(n_value) {
  power_function("gw32", "gw38", "binary_insulin", "binary_EMERGE", n_value)
})
unified_power_results_df <- as.data.frame(t(unified_power_results)) 
colnames(unified_power_results_df) <- c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS")
unified_power_results_df$sample_size <- sample_vector
